---
title: "obAnalytics Guide"
author: "[Phil Nocturne](http://parasec.net)"
date: "`r Sys.Date()`"
output: 
#  rmarkdown::html_vignette:
  html_document:
#    theme: journal 
    toc: true
    highlight: espresso 
  pdf_document:
    toc: true
    highlight: espresso
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Microstructure2 Guide}
  \usepackage[utf8]{inputenc}
---

```{r, include=FALSE}
options(digits.secs=3)
library(obAnalytics)
#knitr::opts_chunk$set(dpi=100, fig.width=8, fig.height=6, results="hide") 
knitr::opts_chunk$set(dpi=100, fig.width=10, fig.height=6, results="hide")
```

# Overview 

_obAnalytics_ is an R package intended for visualisation and analytics of limit
order data. 

This guide is structured as an end-to-end walk-through and is intended to 
demonstrate the main features and functionality of the package.

## Recommended environment settings

Due to the large number of columns in the example data, it is recommended to
set the display width to make the most use of the display. It is also
recommended to set _digits.secs=3_ and _scipen=999_ in order to display
timestamps and fractions nicely. This can be achieved as follows:

```{r}
max.cols <- Sys.getenv("COLUMNS")
options(width=if(max.cols != "") max.cols else 80, scipen=999, digits.secs=3)
```

# Loading data

The main focus of this package is reconstruction of a limit order book. The 
_processData_ function will perform data processing based on a supplied CSV 
file, the format of which is defined in the _Expected csv schema_ section.

The data processing consists of a number of stages:


1. Cleaning of duplicate and erroneous data.

2.  Identification of sequential event relationships.

3.  Inference of trade events via order-matching.

4.  Inference of order types (limit vs market).

5.  Construction of volume by price level series.

6.  Construction of order book summary statistics.


Limit order events are related to one another by volume deltas (the change in 
volume for a limit order). To simulate a matching-engine, and thus determine 
directional trade data, volume deltas from both sides of the limit order book 
are ordered by time, yielding a sequence alignment problem, to which the the 
Needleman-Wunsch algorithm has been applied.

```{r, eval=F}
# load and process example csv data from the package inst/extdata directory.
csv.file <- system.file("extdata", "orders.csv.xz", package="obAnalytics")
lob.data <- processData(csv.file)
```

## Expected csv schema 

The CSV file is expected to contain 7 columns:

--------------------------------------------------------------------------------
           Column name Description
---------------------- ---------------------------------------------------------
                __id__ Numeric limit order unique identifier.

         __timestamp__ Time in milliseconds when event received locally.

__exchange.timestamp__ Time in milliseconds when order first created on the 
                       exchange.

             __price__ Price level of order event.

            __volume__ Remaining order volume.

            __action__ Event action describes the limit order lifecycle. One 
                       of: __created__, __modified__, __deleted__.

         __direction__ Side of order book. On of: __bid__ or __ask__.
--------------------------------------------------------------------------------
Table: Expected CSV schema.

## Preprocessed example data

For illustrative purposes, the package contains a sample of preprocessed data.
The data, taken from the [Bitstamp](https://www.bitstamp.net/websocket) 
(bitcoin) exchange on 2015-05-01, consists of 50,393 limit order events and 482 
trades occuring from midnight up until ~5am.

The sample data, which has been previously processed by the _processData_ 
function, may be attached to the environment with the _data()_ function:

```{r}
data(lob.data)
```

The _lob.data_ object is a list containing four data.frames.

--------------------------------------------------------------------------------
       data.frame Summary
----------------- --------------------------------------------------------------
       __events__ Limit order events.

       __trades__ Inferred trades (executions).

        __depth__ Order book price level depth through time.

__depth.summary__ Limit order book summary statistics.
--------------------------------------------------------------------------------
Table: lob.data summary.

The contents of which are briefly discussed in the following sections.

### Events

The _events_ data.frame contains the lifecycle of limit orders and makes up the
core data of the obAnalytics package. Each row corresponds to a single limit
order action, of which three types are possible.

--------------------------------------------------------------------------------
 Event action Meaning
------------- ------------------------------------------------------------------
  __created__ The order is created with a specified amount of volume and a limit 
              price.

  __changed__ The order has been partially filled. On each modification, the 
              remaining volume will decrease. 

  __deleted__ The order may be deleted at the request of the trader or, in the 
              event that the order has been completely filled, deleted by the 
              exchange. An order deleted by the exchange as a result of being 
              filled will have 0 remaining volume at time of deletion.
--------------------------------------------------------------------------------
Table: Possible limit order actions.

In addition to the event _action_ type, a row consists of a number of 
attributes relating to the lifecycle of a limit order.

--------------------------------------------------------------------------------
             Attribute Meaning 
---------------------- ---------------------------------------------------------
          __event.id__ Event Id.

                __id__ Limit Order Id.

         __timestamp__ Local event timestamp (local time the event was 
                       observed).

__exchange.timestamp__ Exchange order creation time.

             __price__ Limit order price level.

            __volume__ Remaining limit order volume.

            __action__ Event action: _created_, _changed_, _deleted_. (as 
                       described above).

         __direction__ Order book side: _bid_, _ask_.

              __fill__ For _changed_ or _deleted_ events, indicates the change 
                       in volume between this event and the last.

    __matching.event__ Matching _event.id_ if this event is part of a trade. NA
                       otherwise.

              __type__ Limit order type (see Event types below.)

__aggressiveness.bps__ The distance of the order from the edge of the book in 
                       Basis Points (BPS). If an order is placed exactly at the 
                       best bid/ask queue, this value will be 0. If placed 
                       behind the best bid/ask, the value will be negative. A 
                       positive value is indicative of a _innovative_ order: The 
                       order was placed inside the bid/ask spread, which would 
                       result in the change to the market midprice.
--------------------------------------------------------------------------------
Table: Limit order event attributes.

An individual limit order (referenced by the _id_ attribute) may be of six 
different types, all of which have been classified by onAnalytics.

--------------------------------------------------------------------------------
 Limit order type Meaning
----------------- --------------------------------------------------------------
      __unknown__ It was not possible to infer the order type given the 
                  available data. 

__flashed-limit__ Order was created then subsequently deleted. 96% of example 
                  data. These types of orders are also referred to as _fleeting
                  orders_ in the literature.

__resting-limit__ Order was created and left in order book indefinitely until 
                  filled.

 __market-limit__ Order was partially filled before landing in the order book at 
                  it’s limit price. This may happen when the limit order crosses
                  the book because, in the case of a bid order, it's price is >=
                  the current best ask. However there is not enough volume 
                  between the current best ask and the order limit price to fill
                  the order's volume completely.

       __market__ Order was completely filled and did not come to rest in the 
                  order book. Similarly to a _market-limit_, the _market_ order
                  crosses the order book. However, it's volume is filled before
                  reaching it's limit price. Both market-limit and market orders
                  are referred to as _marketable_ limit orders in the 
                  literature.

       __pacman__ A limit-price modified _in situ_ (exchange algorithmic order).
                  The example data contains a number of these order types. They
                  occur when a limit order's _price_ attribute is _updated_. In 
                  the example data, this occurs from a special order type 
                  offered by the exchange which, in the case of a bid, will 
                  _peg_ the limit price to the best ask once per second until 
                  the order has been filled.
--------------------------------------------------------------------------------
Table: Order types.

The following table demonstrates a small snapshot (1 second) of event data. 
Some of the attributes have been omitted or renamed for readability.

```{r}
one.sec <- with(lob.data, {
  events[events$timestamp >= as.POSIXct("2015-05-01 04:55:10", tz="UTC") & 
         events$timestamp <= as.POSIXct("2015-05-01 04:55:11", tz="UTC"),  ]
})
one.sec$volume <- one.sec$volume*10^-8
one.sec$fill <- one.sec$fill*10^-8
one.sec$aggressiveness.bps <- round(one.sec$aggressiveness.bps, 2)
one.sec <- one.sec[, c("event.id", "id", "price", "volume", "action", 
    "direction", "fill", "matching.event", "type", "aggressiveness.bps")]
colnames(one.sec) <- c(c("event.id", "id", "price", "vol", "action", "dir", 
    "fill", "match", "type", "agg"))
print(one.sec, row.names=F)
```

```{r, echo=F, results="markup"}
knitr::kable(one.sec, row.names=F)
```

### Trades

The package automatically infers execution/trade events from the provided limit
order data.

The trades data.frame contains a log of all executions ordered by local 
timestamp. 

In addition to the usual timestamp, price and volume information, each row also
contains the trade direction (buyer or seller initiated) and maker/taker limit 
order ids. 

The maker/taker event and limit order ids can be used to group trades into 
market impacts - An example of which will be demonstrated later in this guide.

```{r}
trades.ex <- tail(lob.data$trades, 10)
trades.ex$volume <- round(trades.ex$volume*10^-8, 2)
print(trades.ex, row.names=F)
```

```{r example trades, echo=F, results="markup"}
knitr::kable(trades.ex, digits=2, row.names=F)
```

Each row, representing a single trade, consists of the following attributes:

--------------------------------------------------------------------------------
          Attribte Meaning
------------------ -------------------------------------------------------------
     __timestamp__ Local event timestamp.

         __price__ Price at which the trade occured.

        __volume__ Amount of traded volume.

     __direction__ The trade direction: _buy_ or _sell_.

__maker.event.id__ Corresponding market making event id in _events_ data.frame.

__taker.event.id__ Corresponding market taking event id in _events_ data.frame.

         __maker__ Id of the market making limit order in _events_ data.frame.

         __taker__ Id of the market taking limit order in _events_ data.frame.
--------------------------------------------------------------------------------
Table: Trade data attributes.


# Visualisation

lala

## Order book shape

The purpose of the cumulative volume graph is to quickly identify the shape of 
the limit order book for the given point in time. The "shape" is defined as the 
cumulative volume available at each price level, starting at the best bid/ask. 

Using this shape, it is possible to visually summarise order book imbalance and
market depth. 

```{r} 
# get a limit order book for a specific point in time, limited to +- 150bps
# above/below best bid/ask price.
lob <- orderBook(lob.data$events, 
    tp=as.POSIXct("2015-05-01 04:38:17.429", tz="UTC"), bps.range=150)

# visualise the order book liquidity.
plotCurrentDepth(lob, volume.scale=10^-8)
```

In the figure above, an order book has been reconstructed with the _orderBook_
function for a specific point in time. The visualisation produced with the
_plotCurrentDepth_ function depicts a number of order book features. Firstly,
the embedded bar chart at the bottom of the plot shows the amount of volume 
available at specific price levels ranging from the _bid_ side on the left 
(blue) through to the _ask_ side (red) on the right. Secondly, the blue and red
lines show the _cumulative_ volume of the bar chart for the bid and ask sides of
the order book respectively. Finally, the two subtle vertical lines at price 
points \$234 and \$238 show the position of the top 1% largest limit orders.


## Price level volume

All lob.data:

```{r}
# plot all lob.data price level volume between $247 and $245 and overlay the 
# market midprice.
spread <- getSpread(lob.data$depth.summary)
plotPriceLevels(lob.data$depth, spread, price.from=227, price.to=245, 
    volume.scale=10^-8, col.bias=0.25, show.mp=T)
```


plot all depth levels, rescaling the volume by 10^-8.
produce 2 plots side-by-side: second plot contains depth levels with > 50 units of volume.

```{r}
spread <- with(lob.data, getSpread(depth.summary))
p1 <- with(lob.data, plotPriceLevels(depth, spread, col.bias=0.1, volume.scale=10^-8))
p2 <- with(lob.data, plotPriceLevels(depth, spread, col.bias=0.1, volume.scale=10^-8, volume.from=50))
library(grid)
pushViewport(viewport(layout=grid.layout(1, 2)))
print(p1, vp=viewport(layout.pos.row=1, layout.pos.col=1))
print(p2, vp=viewport(layout.pos.row=1, layout.pos.col=2))
```

with trades:

```{r}
# plot 1 hour of trades centred around the bid/ask spread. 
plotPriceLevels(lob.data$depth, trades=lob.data$trades, 
    price.from=236, price.to=237.75, volume.scale=10^-8, col.bias=0.2,
    start.time=as.POSIXct("2015-05-01 01:00:00.000", tz="UTC"),
    end.time=as.POSIXct("2015-05-01 02:00:00.000", tz="UTC"))
```

30 minute zoom:

```{r}
# zoom in to 30 minutes of bid/ask quotes.
spread <- getSpread(lob.data$depth.summary)
plotPriceLevels(lob.data$depth, spread, price.from=235.25, price.to=237,
    start.time=as.POSIXct("2015-05-01 00:45:00.000", tz="UTC"), 
    end.time=as.POSIXct("2015-05-01 01:15:00.000", tz="UTC"), 
    volume.scale=10^-8, col.bias=0.5, show.mp=F)
```

4 minute zoom:

```{r}
# zoom in to 4 minutes of bid/ask quotes.
spread <- getSpread(lob.data$depth.summary)
plotPriceLevels(lob.data$depth, spread, price.from=235.90, price.to=236.25,
    start.time=as.POSIXct("2015-05-01 00:55:00.000", tz="UTC"), 
    end.time=as.POSIXct("2015-05-01 00:59:00.000", tz="UTC"), 
    volume.scale=10^-8, col.bias=0.5, show.mp=F)
```

individual market paricipant:

```{r}
# 
spread <- getSpread(lob.data$depth.summary)
plotPriceLevels(lob.data$depth, spread, price.from=232.5, price.to=237.5,
    volume.scale=10^-8, col.bias=1, show.mp=T,
    end.time=as.POSIXct("2015-05-01 01:30:00.000", tz="UTC"),
    volume.from=8.59, volume.to=8.72)
```

```{r}
#
plotPriceLevels(lob.data$depth, price.from=235.65, price.to=237.65,
    volume.scale=10^-8, col.bias=1,
    start.time=as.POSIXct("2015-05-01 01:00:00.000", tz="UTC"),
    end.time=as.POSIXct("2015-05-01 03:00:00.000", tz="UTC"),
    volume.from=3.63, volume.to=3.83)
```

The available volume at each price level is colour coded according to the range of volume at all
price levels. The colour coding follows the visible spectrum, such that larger amounts of volume
appear "hotter" than smaller amounts, where cold = blue, hot = red.
Since the distribution of limit order size exponentially decays, it can be difficult to visually differentiate:
most values will appear to be blue. The function provides price, volume and a colour bias
range to overcome this.

## Liquidity

liquidity...

```{r}
plotVolumePercentiles(lob.data$depth.summary, volume.scale=10^-8, perc.line=F, start.time=as.POSIXct("2015-05-01 01:00:00.000", tz="UTC"),
    end.time=as.POSIXct("2015-05-01 04:00:00.000", tz="UTC"))
``` 

another...
```{r}
# visualise 15 minutes of order book liquidity.
# data will be aggregated to second-by-second resolution.
plotVolumePercentiles(lob.data$depth.summary,
start.time=as.POSIXct("2015-05-01 04:30:00.000", tz="UTC"),
end.time=as.POSIXct("2015-05-01 04:35:00.000", tz="UTC"),
volume.scale=10^-8)
```

## Fleeting orders

```{r}
plotVolumeMap(lob.data$events, volume.scale=10^-8, log.scale = T)
```

```{r}
plotVolumeMap(lob.data$events, volume.scale=10^-8, volume.from=3.5, volume.to=4)
```

```{r}
plotVolumeMap(lob.data$events, volume.scale=10^-8, volume.from=8.59, volume.to=8.72)
```

# Analysis

lala.

## Order book reconstruction

lala.

```{r}
tp <- as.POSIXct("2015-05-01 04:25:15.342", tz="UTC")
ob <- orderBook(lob.data$events, max.levels=10)
print(ob)
```

```{r example order book, echo=F, results="markup"}
tp <- as.POSIXct("2015-05-01 04:25:15.342", tz="UTC")
ob <- orderBook(lob.data$events, max.levels=10)
with(ob, {
  asks$liquidity <- asks$liquidity*10^-8
  bids$liquidity <- bids$liquidity*10^-8
  cols <- c("id", "timestamp", "liquidity", "price")
  knitr::kable(cbind(bids[, cols], asks[order(asks$liquidity), rev(cols)]), 
      row.names=F, align=c("r","r","r","r","l","l","l","l"), digits=2)
})
```

## Market impacts

### all impacts
lala.

```{r}
impacts <- tradeImpacts(lob.data$trades)
impacts <- impacts[impacts$dir == "sell", ]
bps <- 10000 * (impacts$max.price - impacts$min.price) / impacts$max.price
types <- with(lob.data, events[match(impacts$id, events$id), ]$type)
impacts <- cbind(impacts, type=types, bps)
head(impacts[order(-impacts$bps), ], 10)
```

```{r impacts example, echo=F, results="markup"}
impacts <- tradeImpacts(lob.data$trades)
impacts <- impacts[impacts$dir == "sell", ]
impacts <- impacts[, c("id", "max.price", "min.price", "vwap", "hits", "vol", 
    "end.time")]
impacts$vol <- impacts$vol*10^-8
bps <- 10000 * (impacts$max.price - impacts$min.price) / impacts$max.price
types <- with(lob.data, events[match(impacts$id, events$id), ]$type)
impacts <- cbind(impacts, type=types, bps)
knitr::kable(head(impacts[order(-impacts$bps), ], 10), row.names=F, digits=2)
```

### individual impact
lala.

```{r}
impact <- with(lob.data, trades[trades$taker == 65596324, 
    c("timestamp", "price", "volume", "maker")])
makers <- with(lob.data, events[match(impact$maker, events$id), ])
makers <- makers[makers$action == "created", 
    c("id", "timestamp", "aggressiveness.bps")]
impact <- cbind(impact, maker=makers[match(impact$maker, makers$id), 
    c("timestamp", "aggressiveness.bps")])
age <- impact$timestamp - impact$maker.timestamp
impact <-  cbind(impact[!is.na(age), c("timestamp", "price", "volume", 
    "maker.aggressiveness.bps")], age[!is.na(age)])
colnames(impact) <- c("timestamp", "price", "volume", "maker.agg", "age")
impact$volume <- impact$volume*10^-8
print(impact)
```

```{r impact example, echo=F, results="markup"}
impact <- with(lob.data, trades[trades$taker == 65596324,
    c("timestamp", "price", "volume", "maker")])
makers <- with(lob.data, events[match(impact$maker, events$id), ])
makers <- makers[makers$action == "created",
    c("id", "timestamp", "aggressiveness.bps")]
impact <- cbind(impact, maker=makers[match(impact$maker, makers$id),
    c("timestamp", "aggressiveness.bps")])
age <- impact$timestamp - impact$maker.timestamp
impact <-  cbind(impact[!is.na(age), c("timestamp", "price", "volume", 
    "maker.aggressiveness.bps")], age[!is.na(age)])
colnames(impact) <- c("timestamp", "price", "volume", "maker.agg", "age")
impact$volume <- impact$volume*10^-8
knitr::kable(impact, row.names=F, digits=2)
```



